Are Greener Spaces Less Safe Places?

### 2024-04-19 other/datasheet/datasheet_template.qmd
Commit

---
title: "Datasheet for 'A dataset'"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: "`r Sys.time()`"
date-format: "02 April 2024"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

>>>>

---
title: "Datasheet for 'Are Greener Spaces, Safer Places?'"
author: 
  - Jena Shah
thanks: "Code and data are available at: https://github.com/shahjena/greenspace_safeplace.git."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

### 2024-04-19 paper/paper.qmd
Commit

---
title: "Are Greener Spaces, Safer Places?"
subtitle: "Investigating the Link Between Green Spaces and Crime in Toronto"
author: 
  - Jena Shah
thanks: "Code and data are available at: https://github.com/shahjena/greenspace_safeplace.git."
date: today
date-format: long
abstract: "Abstract TBD."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

TBD

# Data {#sec-data}

TBD

@@

TBD

## Model set-up

TBD

## Model justification

TBD

@@

TBD

# Discussion

## First discussion point {#sec-first-point}

TBD

## Second discussion point

TBD

## Third discussion point

TBD

## Weaknesses and next steps

TBD

@@

\appendix

# Appendix {-}

# Datasheet

TBD

# Model details {#sec-model-details}

# Posterior predictive check

TBD

# Diagnostics

TBD

@@

@citeR

>>>>

---
title: "Are Greener Spaces Less Safe Places?"
subtitle: "Exploring the Surprising Relationship Between Urban Green Spaces and Crime Rates in Toronto"
author: 
  - Jena Shah
thanks: "Code and data are available at: https://github.com/shahjena/greenspace_safeplace.git."
toc: true
date: today
date-format: long
abstract: "This study examines the relationship between green spaces and crime rates in Toronto's neighborhoods using gamma regression models. The analysis reveals a positive association between the amount of green spaces and crime rates, indicating that neighborhoods with more green spaces tend to have higher crime rates. This finding emphasizes the relation between urban green spaces and public safety, suggesting that the presence of green spaces may present challenges for crime prevention and urban planning. Understanding these dynamics can help the government and urban planners in their efforts to create safer and more sustainable cities."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(marginaleffects)
library(arrow)
library(rstanarm)
library(ggplot2)
```
\newpage

@@

In urban environments, the presence of green spaces has long been
associated with a variety of benefits: from improved mental health to
enhanced community cohesion. However, the relationship between green
spaces and crime rates is a complex one and has been a subject of debate
in urban planning and crime prevention. There are plenty of studies
suggesting that green spaces may act as a deterrent to crime through
promoting community interaction and physical activity. However, there
are also others who raise concerns that these areas may attract criminal
activity. Understanding this relationship is crucial for creating safer
and more livable cities.

National Geographic has summarized studies indicating that urban areas
with more green spaces tend to have lower crime rates. For example, in
Flint, Michigan - a city that experienced a decrease in crime rates from
2012 to 2022 - the implementation of green spaces was correlated with
this positive change. Similarly, places like Ohio and Philadelphia have
seen recommendations to increase green spaces in areas with high crime
rates as a potential strategy to reduce crime [@burtka2023]. Not only
that, but there is further research backing up the results discussed by
National Geographic. A research paper from the Multidisciplinary Digital
Publishing Institute (MDPI) conducted a comprehensive literature review
to investigate the impact of green spaces on violent crime in urban
settings. Their review included over 30000 potential paper titles and
they ultimately selected 45 papers for inclusion. After their analysis,
their findings and results suggested that green spaces play a
significant role in reducing violent crime. The mechanisms proposed for
this impact include social interaction, social recreation, community
perception, nature-induced stress reduction, climate change, and many
more [@shepley2019].

Unlike the two studies mentioned, this paper will represent the controversial side's contribution to the
current understanding of urban dynamics by focusing on the relationship
between the prevalence of green spaces and crime rates within Toronto's
urban neighbourhoods. By utilizing a rich and expansive dataset along
with a sophisticated statistical model, we conduct a thorough analysis
to unravel the complexities of this relationship. In doing so, the
findings not only focus on the interconnections between green spaces and
crime rates, but also offer insights that can significantly inform the
formulation of urban planning strategies for crime prevention. The
estimand for this study is to quantify the effect of an increase in
green spaces on crime rates in urban neighbourhoods - while controlling
for relevant factors.

The paper is structured into four main sections. First, I provide a
detailed overview of the data utilized in this study, including its
sources and any pre-processing/cleaning steps. I also describe the
methodology used for the analysis, including the statistical figures
made to explore the relationship between green spaces and crime rates.
After that, I present the modeling approach used in this study -
detailing the statistical techniques applied to analyze the data. This
section includes a discussion of the model setup and justification for
the chosen modeling approach. Following that, I present the results of
my analysis, highlighting the key findings regarding the relationship
between green spaces and crime rates in urban neighbourhoods. I use
visualizations such as graphs and tables to illustrate these results
effectively. Finally, the implications of my findings are discussed for
urban planning and crime prevention. I explore how these results can
better the policies which are aimed at creating safer and more
sustainable urban environments. I also take a look at the weaknesses of
my study which could bias the results. Throughout the paper, I utilized
the R programming language and its packages which are cited in the
references.

```{r}
#| label: fig-one
#| fig-cap: Distribution of Green.Spaces and TOTAL_CRIMES.
#| echo: false
#| warning: false
#| message: false
#| eval: true
cleaned_data <- read.csv("C:/Users/shahj/OneDrive/Documents/greencrimes/data/cleaned/cleaned_data.csv")
total_2023 <- cleaned_data$TOTAL_CRIMES
green_spaces_2023 <- cleaned_data$Green.Spaces

par(mfrow = c(1, 2))

hist(total_2023, main = "Total Crimes per 100000, 2023", xlab = "Rate")
hist(green_spaces_2023, main = "Green Spaces 2023", xlab = "Percentage")

par(mfrow = c(1, 1))

```

# Data

## Data Collection and Cleaning

The datasets for this project were obtained from the City of Toronto's
open data portal. There were many other sources that could provide
similar data but they were much older, hence it was more appropriate to
use the data from Open Data Toronto's portal. The first dataset,
"Neighbourhood Crime Rates," contains all neighbourhoods in Toronto and gives the rates (out of 100000) of nine different types of crime from the years 2014 to 2023. Example of crimes include, Assault, Auto Theft, Break and Enter, Robbery, Theft Over \$5000, and more. 
The second dataset, "Wellbeing Toronto Environment," contains information on green spaces and environmental factors. This dataset was provided by Toronto Public
Health, Toronto Parks, Forestry and Recreation, the Federal Ministry of
the Environment, and Toronto Water. It includes data on Green Rebate
Programs, Pollutant Carcinogenic TEP Score, Pollutant Non-Carcinogenic
TEP Score, and Pollutants Released to Air, among other variables.

The datasets were combined using the neighbourhood ID (HOOD_ID) and
filtered for the variables of interest, which include the neighbourhood
ID, neighbourhood name (AREA_NAME), and crime rates for Assault, Auto
Theft, Break and Enter, Robbery, and Theft Over for the year 2023.
Additionally, the dataset includes information on Green Spaces for each
neighbourhood. Since the data of crime across Toronto did not provide a
total amount of crimes per neighbourhood and only focused on the
different types of crime and their rates, I created a new variable which
combined all the different crimes and merged it into a variable called
TOTAL_CRIMES which provides the total amount of crimes per 100000 people
in a neighbourhood.

The decision to focus on the year 2023 out of all the years from 2014 to
2023 was made to use the most recent data available. In the crimes data,
there were a total of 9 types of crimes used but this paper focuses on
using the top 5 crimes in Toronto in the year of 2023 according to
Vilkhov Law. They were selected for analysis and include Assault, Auto
Theft, Break and Enter, Robbery, and Theft Over [@vilkhovlaw]. Other
years and additional crimes were available in the raw crime data, but
for the purposes of this study, only the year 2023 and the selected top
5 crimes were considered.

Overall, the data collection and pre-processing steps involved
obtaining, combining, and filtering the datasets to create a
consolidated dataset for analysis. These steps ensured that the data
used in the study were relevant, accurate, and suitable for
investigating the relationship between green spaces and crime rates in
Toronto's neighbourhoods.

## Data Variables

The dataset for this study comprises four key variables that are central
to the analysis. These variables are properly descibed below.

1.  HOOD_ID (Neighbourhood ID): This variable serves as a unique
    identifier for each neighborhood in Toronto, allowing us to
    differentiate between different areas in our analysis.

2.  AREA_NAME (Neighbourhood Name): This variable provides the name of
    each neighborhood, providing context and a more intuitive
    understanding of the data.

3.  Green.Spaces (Amount of Green Spaces): This variable quantifies the
    amount of green spaces in each neighborhood, measured in square
    kilometers. Green spaces include parks, forests, and other areas
    covered by vegetation.

4.  TOTAL_CRIMES (Total Crimes per Neighbourhood per 100000 People):
    This variable represents the total number of reported crimes per
    neighborhood, standardized to a rate per 100000 people. It provides
    a measure of the overall crime burden in each neighborhood,
    accounting for differences in population size.

These variables were selected based on their relevance to the study's
objective of examining the relationship between green spaces and crime
rates. The inclusion of the neighborhood ID and name allows us to
identify and differentiate between neighborhoods, while the Green.Spaces
variable provides a measure of the presence and extent of green spaces
in each area. The TOTAL_CRIMES variable provides a measure of the
overall crime rates, allowing us to assess the impact of green spaces on
crime in different neighborhoods. @fig-one shows the different variables
and we can see the right-skew in the crimes total, which can help us
decide what regression model to use in the next step.

@@

## Model Set-up

This study was done to investigate the relationship between the abundance of green
spaces and crime rates in urban neighborhoods in Toronto by using a
gamma regression model. This approach is suitable for analyzing count
data with a skewed distribution, such as crime rates, where the response
variable is non-negative and continuous.

### Variables

The key variables used in the model are:

-   Dependent Variable: TOTAL_CRIMES (total crimes per neighborhood per
    100000 people).

-   Independent Variable: Green.Spaces (amount of green spaces in square
    kilometers).

### Model Details

The gamma regression model is specified as follows:

$log(E(Y)) = \beta_0 + \beta_1 \times \mbox{Green.Spaces}$

Where:

-   (E(Y)) is the expected value of the dependent variable
    (TOTAL_CRIMES).

-   $\beta_0$ represents the intercept term.

-   $\beta_1$ is the coefficient associated with the Green.Spaces
    variable, indicating the effect of green spaces on crime rates.

### Assumptions

The gamma regression model assumes that the dependent variable follows a
gamma distribution. We also assume that the relationship between green
spaces and crime rates is linear, meaning that a one-unit increase
in green spaces is associated with a constant percentage change in crime
rates.

## Model Justification

The choice of a gamma regression model is appropriate for this study
because it is specifically designed for analyzing skewed count data. We noticed the right-skew in the histograms of green spaces and total crimes in @fig-one. 
By using a gamma regression approach, we can account for the skewed nature of crime rate data and obtain more reliable estimates of the
relationship between green spaces and crime rates.

Overall, the gamma regression model is well-suited for analyzing the
relationship between green spaces and crime rates in Toronto's
neighborhoods, providing a statistically interpretable
approach to examining this important issue.

@@

In the results, I found that green spaces have a statistically significant relationship with crime rates in Toronto's neighborhoods. The gamma regression model showed that as the amount of green spaces in a neighborhood increases, the total crimes per 100000 people also tend to increase. Specifically, for every unit increase in green spaces (measured in square kilometers), there was a corresponding increase in total crimes.

@tab-one presents the results of the explanatory model for crime rates based on green spaces. The intercept and coefficient for green spaces are both statistically significant, indicating that green spaces have a significant impact on crime rates. The model had 122 observations, and the log-likelihood was -943.227. The model's predictive performance was assessed using ELPD, LOOIC, WAIC, and RMSE, indicating a good fit of the model to the data.

@fig-one illustrates this relationship with a scatterplot. It shows a concentration of data points at low to medium crime rates and low levels of green spaces. As the amount of green spaces increases, there are much fewer data points and they are more scattered, indicating higher crime rates in neighborhoods with more green spaces. The line of best fit in the scatterplot shows a nearly straight line, yet still highlighting the positive relationship between green spaces and crime rates.

@fig-two enhances this by presenting a line graph that shows the uncertainty shadow of the line of best fit. It demonstrates that the shadow is small at low to medium crime rates and low levels of green spaces, indicating higher certainty in the relationship. However, as the amount of green spaces increases, the shadow expands a lot, indicating more uncertainty in the relationship.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

cleaned_par <- read_parquet("C:/Users/shahj/OneDrive/Documents/greencrimes/data/cleaned/cleaned_data.parquet") %>%
  filter(!is.na(TOTAL_CRIMES) & TOTAL_CRIMES != 0 & !is.na(Green.Spaces) & Green.Spaces != 0)

```

```{r}
#| label: fig-two
#| fig-cap: Crime rate (out of 100000) frequency depending on the area of green spaces (square kilometers)
#| echo: false
#| warning: false
#| message: false
#| eval: true

ggplot(cleaned_par, aes(x = Green.Spaces, y = TOTAL_CRIMES)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Total Area of Green Spaces in square kilometers",
    y = "Total Crimes per 100000",
    title = "Relationship between Green Spaces and Total Crimes"
  ) +
  theme_minimal()

```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

my_model <-
  readRDS(file = here::here("C:/Users/shahj/OneDrive/Documents/greencrimes/models/model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-one
#| tbl-cap: "Explanatory models of the crime rates based on the green spaces"
#| warning: false

modelsummary::modelsummary(
  list(
    "Model" = my_model
  ),
  statistic = "mad",
  fmt = 2 
)
```

```{r}
#| echo: false
#| eval: true
#| label: fig-three
#| fig-cap: "The number of crimes predicted by the model for the area of green spaces per neighbourhood"
#| warning: false

plot_predictions(my_model, condition = "Green.Spaces") +
  labs(x = "Green Spaces (square kilometers)",
       y = "Crime Rate (per 100000)") +
  theme_classic()
```

# Discussion
In this study, I analyzed the relationship between green spaces and crime rates in Toronto's neighborhoods using gamma regression models. The results suggest a positive relationship between the amount of green spaces and crime rates, indicating that areas with more green spaces tend to have higher crime rates. However, it is important to note that this relationship does not imply causation.

One of the key insights gained from this study is the potential impact of green spaces on crime rates in urban areas. The findings suggest that neighborhoods with more green spaces may face unique challenges in terms of crime prevention and urban planning. This knowledge can inform policymakers and urban planners in their efforts to create safer and more sustainable cities.

## Weaknesses and Next Steps

Several limitations in this study should be acknowledged. Firstly, the green space data used in this analysis is from 2011, which may not accurately reflect the current green space distribution in Toronto. Additionally, the process of reducing the number of neighborhoods from 158 to 138 due to the removal the NA and 0 values could introduce bias into the analysis (so a total of 20 nighbourhoods were not considered in this study).

To deepen the understanding of relationships between green spaces and crime rates, future research should focus on using more recent green space data and exploring other factors that may influence crime rates in urban areas. For instance, economic status, social factor, and other demographics relating to the different neighbourhoods. Additionally, conducting similar studies in other cities instead of Toronto could help to determine the general idea behind these findings.

## Conclusion
In conclusion, this study provides details about the relationship between green spaces and crime rates in urban areas. By understanding the impact of green spaces on crime rates, the policymakers, government and urban planners can work towards creating even more safe and sustainable cities for their residents.

@@

# Appendix

## Datasheet

## Posterior predictive check
This check is a way to assess the goodness-of-fit of a model by comparing its predictions to actual observations. (a) Peaks in @fig-pp check around 1000-1500 indicate areas where the model may not accurately capture the data's variability or patterns. (b) The posterior estimates are more in the middle, indicating greater certainty after adding in the data, whereas prior estimates are more dispersed, reflecting greater uncertainty before observing the data.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-pp
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits and is affected by the data"
#| fig-subcap: ["Posterior prediction check", "Comparing posterior with prior"]

pp_check(my_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(my_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics
The diagnostics help assess whether the model has sampled from the posterior distribution effectively. In @fig-mcmc (a) The trace plot shows that chain 4 appears to have higher variability than the others. (b) The Rhat plot displays values around 1 indicating convergence of the chains.
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-mcmc
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(my_model, "trace")

plot(my_model, "rhat")
```

@@

@citeR @citeODT @burtka2023 @shepley2019 @vilkhovlaw
